---
title: 'The Pilots: Delay Decoder Final Report'
author: Sophia Chavez, Elif Turgut, Joseph Nam, Justin Henriquez, Arjun Sharma, Lance
  Santana
date: "2025-06-08"
output:
  html_document: default
  pdf_document: default
subtitle: "STAT 167 - Introduction to Data Science"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Description

Hundreds of millions of people travel each year, and hundreds of
millions of people have delayed flights. What if we could minimize the
number of passengers affected by delayed flights? Throughout this
project, we will investigate correlations between flight delays and
conditions such as weather, plane type, airline, etc. with the goal of
assisting travelers in booking their flights with the best probability
of avoiding a delayed flight.

We will be using the anyflights dataset which includes information on
all flights departing from airports in a given year including the
“nycflights13” data. Specifically, extracting all 2013 flights departing
from America's top ten busiest airports; ATL, LAX, DFW, DEN, ORD, MCO,
JFK, LAS, CLT, MIA.

Assumptions: To make sure our comparisons and conclusions are accurate,
we will assume that there is consistency in how certain variables are
measured across the different datasets, most importantly for how the
weather factors and delay times are recorded.

## Dataset

```{r, echo=FALSE, }
library(dplyr)
library(ggplot2)
library(anyflights)
library(nycflights13)
library(patchwork)
library(anyflights)
options(timeout = 800)
  
busiest_flights13_ <- get_flights(c("ATL", "LAX", "DFW","DEN","ORD", "MCO","JFK", "LAS", "CLT", "MIA" ), 2013)

  

  
```

```{r, echo=FALSE, message=FALSE}
glimpse(busiest_flights13_)
  
weather_data <-  get_weather(c("ATL", "LAX", "DFW","DEN","ORD", "MCO","JFK", "LAS", "CLT", "MIA" ), 2013)
glimpse(weather_data)
```

## Month Analysis

```{r, message=FALSE}

busiest_flights13_ <- filter(busiest_flights13_, !is.na(dep_delay), !is.na(arr_delay)) # removing canceled flights


busiest_flights13_grouped <- group_by(busiest_flights13_ , origin, month)

data <- summarize(busiest_flights13_grouped,
          flight_count = n(),
          avg_delay = mean(dep_delay, na.rm = TRUE),
          delay_total = sum(dep_delay>0, na.rm = TRUE),
          avg_distance = mean(distance, na.rm=TRUE),
          prop_delay = delay_total / n())

busiest_flights13_summary <- arrange(data, desc(prop_delay))


ggplot(data = busiest_flights13_summary) +
geom_point(mapping = aes(x = origin, y = prop_delay, color = origin)) +
facet_wrap(~ month, nrow = 2)
```

June, July and December had the most delayed flights regardless of
origin. This tells us that we should pay closer attention to these
months and their weather conditions, plane types, and distance covered
to try and identify what is causing the rise in delayed flights.

## Time Analysis

```{r, echo=FALSE}
# Convert the 'hour' column to parts of the day
flights <- busiest_flights13_
flights$time_of_day <- cut(flights$hour,
                            breaks = c(0, 5, 12, 17, 21, 24), 
                            labels = c("Night", "Morning", "Afternoon", "Evening", "Night"),
                            include.lowest = TRUE, right = FALSE)

flights$time_of_day <- factor(flights$time_of_day, 
                               levels = c("Morning", "Afternoon", "Evening", "Night"))

# Check the results
table(flights$time_of_day)

# Group by time of day and calculate average departure delay
avg_delay_by_time <- flights %>%
  group_by(time_of_day) %>%
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE))


# Bar plot of average departure delay by time of day
ggplot(avg_delay_by_time, aes(x = time_of_day, y = avg_dep_delay)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Average Departure Delay by Time of Day", 
       x = "Time of Day", y = "Average Departure Delay (minutes)")

# ANOVA test to check if time of day affects departure delay
anova_results <- aov(dep_delay ~ time_of_day, data = flights)
summary(anova_results)
```

Ho: Delays do not depend on time of day

Ha: Delays depend on the time of day

We calculated a small p-value, which means that we can reject the null
hypothesis and have evidence that the time of day has a significant
effect on departure delays, which is also seen in the graph.

## Distance Analysis

```{r, echo=FALSE}

ggplot(data = busiest_flights13_summary) +
geom_point(mapping = aes(x = avg_distance, y = prop_delay, color = origin)) +
facet_wrap(~month)

```

The plot above shows that regardless of month and origin, the
distribution follows the same pattern. The amount of delayed flights
peak between 700 and 1000 miles traveled therefore we can state that the
proportion of delayed flights and distance traveled do not have a
positive correlation.

## Weather Analysis

```{r, echo=FALSE, message=FALSE}
  

#cleaning weather data
weather_data <- filter(weather_data, !is.na(temp), !is.na(humid),!is.na(wind_dir),!is.na(wind_speed),!is.na(wind_gust),!is.na(precip),!is.na(pressure),!is.na(visib))



#joining weather and flights
flights_weather <- left_join(
busiest_flights13_grouped, weather_data, by = c("origin" = "origin", "time_hour" = "time_hour",  "month" = "month")) 



flights_weather_grouped <- group_by(flights_weather, month, origin)
flights_weather_summary <- summarize(flights_weather_grouped,
                                    delay_total = sum(dep_delay>0, na.rm = TRUE),
                                    prop_delay = delay_total / n(),
                                    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
                                    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
                                    avg_air_time = mean(air_time, na.rm = TRUE),
                                    avg_temp = mean(temp, na.rm = TRUE),
                                    avg_humid = mean(humid, na.rm = TRUE),
                                    avg_wind_dir = mean(wind_dir, na.rm = TRUE),
                                    avg_wind_speed = mean(wind_speed, na.rm = TRUE),
                                    avg_wind_gust = mean(wind_gust, na.rm = TRUE),
                                    avg_precip = mean(precip, na.rm = TRUE),
                                    avg_pressure = mean(pressure, na.rm = TRUE),
                                    avg_visib = mean(visib, na.rm = TRUE),)

lm.fit <- lm(prop_delay ~  avg_dep_delay + avg_arr_delay + avg_air_time + avg_temp + avg_humid + avg_wind_dir + avg_wind_speed + avg_wind_gust + avg_precip +avg_pressure +avg_visib, flights_weather_summary)
summary(lm.fit)


wind_gust_lm.fit <- lm(prop_delay ~ avg_wind_gust, flights_weather_summary)
summary(wind_gust_lm.fit) #shows small relation 4.1% variance represented by wind gust

```

```{r, echo=FALSE}
flights_weather_summary$residuals <- resid(lm.fit)
flights_weather_summary$predicted_delay <- predict(lm.fit)

ggplot(flights_weather_summary, aes(x = predicted_delay, y = residuals)) +
  geom_point(color = "darkorange", size = 3, alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Predicted Values",
       x = "Predicted Proportion Delay",
       y = "Residuals") +
  theme_minimal()
```

```{r, echo=FALSE}
# Add predicted values to your dataset
flights_weather_summary$predicted_delay <- predict(lm.fit)

# Plot
ggplot(flights_weather_summary, aes(x = prop_delay, y = predicted_delay)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Predicted vs Actual Proportion of Delayed Flights",
       x = "Actual Proportion Delay",
       y = "Predicted Proportion Delay") +
  theme_minimal()
```

```{r, echo=FALSE, message=FALSE}
ggplot(flights_weather_summary, aes(x = avg_humid, y = prop_delay)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(title = "Proportion of Delayed Flights vs. Humidity",
       x = "Average Humidity", y = "Proportion of Delayed Flights")

```

```{r, echo=FALSE, message=FALSE}
ggplot(flights_weather_summary, aes(x = avg_precip, y = prop_delay)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "purple") +
  labs(title = "Proportion of Delayed Flights vs. Precipitation",
       x = "Average Precipitation", y = "Proportion of Delayed Flights")
```

We performed linear regression on the various weather conditions to find
strong relationships between the proportion of delayed flights and
weather conditions. Based on our results, that as the average humidity
and precipitation increases, the proportion of delayed flights
decreases. Additionally, as wind gust increases so does the proportion
of delayed flights.

## Aircraft Analysis

```{r, echo=FALSE}
top_airports<- flights %>%
  count(origin, sort=TRUE) %>%
  slice_max(n,n=10)
busiest_flights13_<- flights %>%
  filter(origin %in% top_airports$origin)
busiest_flights_with_planes<- busiest_flights13_ %>%
  left_join(planes, by="tailnum")
head(busiest_flights_with_planes)

# Correlation between delays and plane features(manufacturer year)
#cor(busiest_flights_with_planes$dep_delay,busiest_flights_with_planes$year.y,use="complete.obs")

filtered_data<-busiest_flights_with_planes %>%
  filter(year.y >= 1990)
ggplot(filtered_data, aes(x=factor(year.y),y=dep_delay))+
  geom_violin(fill="lightblue",alpha=0.7)+
  labs(
    title="Departure Delay vs Plane Manufacture Year",
    x="Manufacture Year of Plane",
    y="Departure Delay"
  )
```

We are looking to see if there is a correlation between delays and plane
features such as the manufacturer year. I filtered the data so that we
would be looking at the most recent years for this analysis. Looking at
the results we can see that in the early manufacture years the departure
delays are high and they gradually decrease as time goes on with a
couple of outliers in 2002 and especially in 2011. But for the most part
as the manufacture year of the plane gets newer the departure delays
decrease.

Limitation: A limitation that I found while working on the Aircraft
Analysis was the sheer amount of data that had to be processed and
analyzed. It would take a really long time to load the data and conduct
any visualizations because of how large the data set was. To fix this
issue I filtered the data to only include the manufacture year from 1990
and up which gives us a higher level of data reliability because it is
comparing planes in recent modern time.

## Origin analysis

Looking at our prop_delayed and origin facet plot, we can see that
specific origin airports consistently have the highest or smallest
proportion of delayed flights. Is this trend statistically significant?

```{r}
lm.fit_carrier_origin <- lm(prop_delay ~   origin, data = busiest_flights13_summary)
summary(lm.fit_carrier_origin)
```

Looking at the results, we can see that origins DEN,DFW,LAS,and ORD have
a significant relationship to the proportion of delayed flights.
Specifically, DEN, DFW, LAS and ORD lead to higher proportions of
delayed flights. What conditions could be contributing to this trend?

```{r, echo=FALSE}


origin_weather_data<-mutate(flights_weather, class = ifelse(origin == "DEN"|origin ==  "DFW"| origin == "LAS"|origin == "ORD",1,0))
origin_weather_data <- group_by(origin_weather_data, class)
flights_weather_summary <- summarize(origin_weather_data,
                                    delay_total = sum(dep_delay>0, na.rm = TRUE),
                                    prop_delay = delay_total / n(),
                                    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
                                    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
                                    avg_air_time = mean(air_time, na.rm = TRUE),
                                    avg_temp = mean(temp, na.rm = TRUE),
                                    avg_humid = mean(humid, na.rm = TRUE),
                                    avg_wind_dir = mean(wind_dir, na.rm = TRUE),
                                    avg_wind_speed = mean(wind_speed, na.rm = TRUE),
                                    avg_wind_gust = mean(wind_gust, na.rm = TRUE),
                                    avg_precip = mean(precip, na.rm = TRUE),
                                    avg_pressure = mean(pressure, na.rm = TRUE),
                                    avg_visib = mean(visib, na.rm = TRUE),)
flights_weather_summary
```

In order to see if weather conditions correlate to the higher proportion
of delayed flights, we split the data set into two classes, one
containing origins that have a statistically significant relationship to
the proportion of delayed flights and the other containing origins
without a significant relationship. Our results show that flights
departing from origins that have a statistically significant
relationship to the proportion of delayed flights also have
significantly higher wind gusts. These results further our weather
analysis results, confirming higher wind gusts and speeds are correlated
to a higher proportion of delayed flights.

## Flight Volume vs. Delay Rates by Month

```{r, echo=FALSE}
#calculates the number of flights that occure each month
monthly_flights <- flights %>% 
  group_by(month) %>% 
  summarise(amt_flights = n())

#calculates the number of flights with delays that occure each month
monthly_delays <- flights %>% 
  mutate(delay = dep_delay > 15) %>% 
  group_by(month) %>% 
  summarise(amt_delay_flights = sum(delay, na.rm = TRUE))

#combines both counts into one dataset
monthly_both <- monthly_flights %>% inner_join(monthly_delays, by = "month")

#graphs the number of flights each month
ggplot(monthly_both, aes(x = factor(month))) +
  geom_col(aes(y = amt_flights, fill = "Total Flights"), position = "dodge") +
  geom_line(aes(y = amt_delay_flights, color = "Delayed Flights", group = 1)) +
  geom_point(aes(y = amt_delay_flights, color = "Delayed Flights", group = 1)) +
  scale_fill_manual(name = "Legend", values = c("Total Flights"= "orange")) + 
  scale_color_manual(name = "Legend", values = c("Delayed Flights"= "red")) + 
  labs(
    title = "Amount of Flights by Month",
    x = "Month",
    y = "Amount of Flights"
  )
```

This analysis was based off the idea that months with higher demand for
travel ie. more flights would result in increased delays. The above plot
demonstrated a mild correlation between the total flights occurring in
each month and the amount of delays. However as the change total number
of flights has a direct effect on the number of delays it is necessary
to standardize the data with z-scores for each month to see a real
relationship.

```{r, echo=FALSE , message=FALSE}
#calculates the number of flights that occure each month
monthly_flights <- flights %>% 
  group_by(month) %>% 
  summarise(amt_flights = n()) %>%
  mutate(z_score_total = scale(amt_flights)[,1])

#calculates the number of flights with delays that occure each month
monthly_delays <- flights %>% 
  mutate(delay = dep_delay > 15) %>% 
  group_by(month) %>% 
  summarise(amt_delay_flights = sum(delay, na.rm = TRUE))

#calculates the percentage of flights with delays that occure each month
monthly_delay_percent <- monthly_flights %>%
  inner_join(monthly_delays, by = "month") %>%
  mutate(
    percentage_delay = 100 * amt_delay_flights / amt_flights, 
    z_score = scale(percentage_delay)[,1]
    )

#graphs the standardized amount of flights each month
plot_1 <- ggplot(monthly_delay_percent, aes(x = factor(month), y = z_score)) +
  geom_col(fill = "red") +
  ylim(-2, 2) +
  labs(
    title = "Percentage of Flights Delayed by Month",
    x = "Month",
    y = "Z-Score of Percent Flights Delayed"
  )

#graphs the standardized amount of delays each month
plot_2 <- ggplot(monthly_delay_percent, aes(x = factor(month), y = z_score_total)) +
  geom_col(fill = "blue") +
  ylim(-2, 2) +
  labs(
    title = "Total Flights by Month",
    x = "Month",
    y = "Z-Score of Total Flights "
  )

plot_1 + plot_2
```

After standardizing the data with z-scores there is a somewhat of a
detectable relationship. Months 1, 2 are negative or 0, and 3-8 are
positive on both graphs, however the variability does not match with
significant increases and drops in total flights, flight 10 sticks out
as a noticeable outlier as flight delay is low negative while total
flights is a high positive. Overall this analysis heavily called into
question a noteable relationship between the two variables.

```{r, echo=FALSE, message=FALSE}
#trains model to perdict delay form the total amount of flights
model <- lm(percentage_delay ~ z_score_total, data = monthly_delay_percent)

#summary of results
summary(model)

#graphs regression line
ggplot(monthly_delay_percent, aes(x = z_score_total, y = percentage_delay)) +
  geom_point(color = "blue", size = 2.5) + 
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Regression Line: % Flight Delay by Flight Amount",
    x = "Z-Score for Total Flights",
    y = "Percentage of Flights Delayed"
  )
```

After performing linear regression and graphing, the results demonstrate
a high p-value of .216 above the significant .05 level therefore we
cannot say that the increase in flight delays is associated with a
increase in flight quantity. The R-Squared value added to this analysis
with the only 15% variability in delay volume being explained by flight
quantity. Overall there seems to be a weak positive trend between the
two variables but it is not statistically significant. Limitations of
this analysis include the focus on monthly trends without accounting for
daily variation or other influencing variables such as weather.
Comparing flight delays by week, time of day, or specific dates might
reveal more consistent and meaningful patterns.

## Carrier Analysis

We are also interested to see if the travel airline or carrier plays a
role flight delays. To do this, we performed a one-way ANOVA test to see
if the mean departure delay times significantly differ by carrier.

```{r,collapse=T, echo=FALSE}
aov_carrier <- aov(dep_delay ~ carrier, data = busiest_flights13_)
summary(aov_carrier)
```

The small p-value (\<2e-16) tells us that there is evidence for a
significant difference in mean departure delay times across the
carriers. We can investigate this further with Tukey's HSD test, which
compares each carrier pair for a significant difference and allows us to
rank the carriers based on which pairs are significantly different from
one another and which ones are not.

```{r,collapse=T , echo=FALSE}
tukey <- TukeyHSD(aov_carrier)

plot(tukey)

# install.packages("agricolae")
library(agricolae)
hsd <- HSD.test(aov_carrier, "carrier", group = TRUE)
hsd$groups
```

The TukeyHSD() function takes each carrier pair and provides a 95%
confidence interval and p-value for the true difference in means. Then,
the HSD.test() function streamlines the data for each pair into the
appropriate ranking, where each group letter represents an individual
rank. For example, carries VX (Virgin America), YV (Mesa Airlines), and
FL (AirTran Airways) are all under group g, which means the three
carriers are not considered significantly different from one another and
are all "tied" for that rank. Groups with 2 letters (i.e. bc) serve as
bridges between two different groups, as they are not significantly
different from either group.

Based on this ranking, Alaska Airlines (AS), US Airways (US), and a tie
between Delta and SkyWest Airlines (DL and OO) are ranked as the top 3
airlines with the shortest delays, making them the safest travel
options. ExpressJet Airlines (EV), Southwest Airlines (WN) and Endeavor
Air (9E) (tie since 9E is group bc but closer to group b), and Frontier
Airlines (F9) are ranked as the top 3 airlines with the longest delays,
which makes them the least ideal airlines to travel with to avoid
delays.

We also ran a logistic regression model that investigates and predicts
the likelihood of each carrier experiencing a delay. To do this, we
converted the departure delay times into a binary variable,
corresponding to a delayed vs not delayed flight. We then turned the
model's results into a barplot of predicted probabilities by predicting
the probability of each flight being delayed, then averaging the
probability across each carrier.

```{r,collapse=T, echo=FALSE}
#busiest_flights13_$dep_delay_bin <- ifelse(busiest_flights13_$dep_delay > 0, 1, 0)

#logit_model <- glm(dep_delay_bin ~ carrier, data = busiest_flights13_, family = "binomial")
#summary(logit_model)

#busiest_flights13_$predicted_prob <- predict(logit_model, type = "response")
#carrier_probs <- busiest_flights13_ %>% 
  #group_by(carrier) %>%
  #summarize(avg_prob = mean(predicted_prob)) # Calculate average predicted probability by carrier

#ggplot(carrier_probs, aes(x = reorder(carrier, -avg_prob), y = avg_prob)) +
 # geom_bar(stat = "identity", fill = "skyblue") +
 # labs(title = "Predicted Probability of Delay by Carrier",
    #   x = "Carrier", y = "Predicted Probability")
```

Based on the plot, Southwest Airlines is the carrier most likely to
experience a flight delay at around 65%, while Alaska Airlines is the
least likely, at just over 20%. In general, most airlines that were
ranked with higher delay times tend to have higher delay probabilities,
while the opposite is true for airlines with lower delay times on
average. This further supports the validity of the ranking from the
Tukey's test and overall suggestions for which airlines to opt for or
avoid.

We are also interested in seeing if airlines are consistent across
different origins in terms of being delayed or on time. We used a
two-way ANOVA procedure to analyze the interaction effects between both
the carrier and origin on the departure delays. The two-way ANOVA also
shows the individual effects of both the origin and carrier, which have
both already been established to have an effect on departure delay
times.

```{r, echo=FALSE}
aov_model2 <- aov(dep_delay ~ carrier * origin, data = busiest_flights13_)
summary(aov_model2)
```

The small p-value (\<2e-16) for the interaction effects (carrier:origin)
tells us there is significant evidence that the effects of the carrier
on flight delays depends on the flight origin, and vice versa. This
means that in general, at least some airlines are likely not consistent
across all origins. This can be verified using the plot below:

```{r,echo=FALSE}
ggplot(data = busiest_flights13_) +
geom_point(mapping = aes(x = carrier, y = dep_delay, color = carrier)) +
facet_wrap(~ origin, nrow = 2) +
labs(x = "Carrier", y = "Departure Delay", title = "Airline Delays by Origin Airport")
```

The plot shows consistencies in airlines like American Airlines (AA) and
Delta Airlines (DL), which ranked higher in delay times, as the spread
of delay times are around the same size across most or all origins.
Airlines like Skywest (OO) and ExpressJet (EV), and Envoy Air (MQ) are
less consistent due to the spread of delay times being more varied
across different origins. This variability support the ANOVA procedure's
result, in addition to the fact that some carriers don't operate at
certain locations, which in turn adds more significance to the
difference in delays across each carrier at each origin.

Carrier analysis conclusion

We found a difference in delay times based on the carrier using an ANOVA
procedure, and we used that information to determine which carriers will
most likely be on time and which ones to potentially avoid by using
Tukey's test. We then followed up with a prediction model using logistic
to regression to predict the odds of a flight being delayed based on the
given carrier, and the results supported our previous findings. Finally,
we conducted a two-way ANOVA procedure to see if the interaction effects
between the carrier and flight origin have significant impacts on flight
delay or if airlines are consistent across different origins. This
procedure also showed a significant difference, which we observed and
confirmed in the above plot by looking at the spread of delays for each
carrier across all origins.

## Conclusions

Our research has identified conditions most and least associated with
delayed flights. We can assist passengers departing from one of the top
10 busiest airports by booking flights with the best probability of
avoiding a delayed flight.

-   Depart early morning

-   If possible fly during January through May, or September through
    November

-   On the most recently manufactured plane

-   On days with low wind gusts

-   If possible, opt for airlines such as Alaska or Delta and avoid
    airlines such as Frontier or Southwest

General Limitation:

A key limitation of our analysis was the use of the anyflights()
function, which downloads data in real time from an external source. Due
to the slow response time and large data volume, we limited our dataset
to only the top 10 busiest U.S. airports. This allowed for more
efficient processing but may have excluded important patterns present at
smaller or regional airports.

## Authors' Contributions

-   Sophia Chavez: Month, distance, and weather analysis. Multiple
    linear regression. Final report setup

-   Elif Turgut: Weather analysis visuals, time of day analysis

-   Joseph Nam: Objective analysis, time of day analysis, plane analysis

-   Justin Henriquez: Background, objective analysis

-   Arjun Sharma: Carrier and origin analysis

-   Lance Santana: Delay vs flight quantity analysis, final slide deck

## Data Availability

R documentation for 'anyflights':
<https://github.com/simonpcouch/anyflights>

nycflights13 package:
<https://cran.r-project.org/web/packages/nycflights13/index.html>
